\chapter{Uposatha}

\section{Pārisuddhi-uposatha (Purity Uposatha)}

\subsection{Pārisuddhi Before Sangha}

Declaring one's purity before the Sangha:

‘Parisuddho ahaṃ bhante, parisuddho'ti maṃ saṅgho dhāretu.’

‘\emph{I, ven. sirs, am quite pure (of offences). May the Saṅgha hold me to be pure.}’

\suttaRef{Vin.I.120–129}

\subsection{Pārisuddhi for Three Bhikkhus}

The Pātimokkha requires at least four bhikkhus. If there are only three bhikkhus then, after the preliminary duties and the general confession, one bhikkhu
chants the \emph{ñatti}:

‘Suṇantu me bhante āyasmantā ajj'uposatho paṇṇaraso, yad'āyasmantānaṃ
pattakallaṃ, mayaṃ aññamaññaṃ pārisuddhi uposathaṃ kareyyāma.’

‘\emph{Let the ven. ones listen to me. Today is an Observance day, which is a
  fifteenth (day of the fortnight). If it seems right to the ven. ones let
  us carry out the Observance with one another by way of entire purity.}’

When it is the 14th day:\\
‘paṇṇaraso’ → ‘cātuddaso’

If the announcing bhikkhu is the most senior:\\
‘bhante’ → ‘āvuso’

Then, starting with the senior bhikkhu:

‘Parisuddho ahaṃ āvuso,\\
parisuddho'ti maṃ dhāretha.’ (×3)\\
‘\emph{I, friends, am quite pure. Understand that I am quite pure.}’

For each of the two junior bhikkhus:\\
‘āvuso’ → ‘bhante’

\subsection{Pārisuddhi for Two Bhikkhus}

Omit the \emph{ñatti}. The senior bhikkhu declares purity first:

‘Parisuddho ahaṃ āvuso, parisuddho'ti maṃ dhārehi.’ (×3)

For the junior:\\
‘āvuso’ → ‘bhante’\\
‘dhārehi’ → ‘dhāretha’

\subsection{Adhiṭṭhānuposatha (For a lone bhikkhu)}

For a bhikkhu staying alone on the Uposatha day. After the preliminary duties,
he then determines:

‘Ajja me uposatho.’\\
‘\emph{Today is an Observance day for me.}’

\section{Sick Bhikkhus}

\subsection{Pārisuddhi}

\textbf{(a)} The sick bhikkhu makes general confession, then:

‘Pārisuddhiṃ dammi, pārisuddhiṃ me hara, pārisuddhiṃ me ārocehi.’

‘\emph{I give my purity. Please convey purity for me (and) declare purity for me.}’

If the sick bhikkhu is the junior:\\
‘hara’ → ‘haratha’\\
‘ārocehi’ → ‘ārocetha’

\clearpage

\textbf{(b)} The sick bhikkhu's (e.g. Uttaro's) purity is conveyed after the
Pātimokkha:

‘Āyasmā bhante ‘uttaro’ bhikkhu gilāno, parisuddho'ti paṭijāni, parisuddho'ti taṃ saṅgho dhāretu.’\\
‘\emph{Ven. sirs, ‘Uttaro Bhikkhu’ who is sick acknowedges that he is pure. May
  the Saṅgha hold him to be pure.}’

If the bhikkhu conveying purity is senior to the sick bhikkhu:

‘Āyasmā bhante \emph{uttaro}’ → ‘\emph{Uttaro} bhante bhikkhu’

\subsection{Sending Consent (Chanda)}

\textbf{(a)} The sick bhikkhu sends his consent to the \emph{saṅghakamma}:

‘Chandaṃ dammi, chandaṃ me hara, chandaṃ me ārocehi.’\\
‘\emph{I offer my consent. May you convey my consent (to the Saṅgha). May you
  declare my consent to them.}’

If the sick bhikkhu is the junior:\\
‘hara’ → ‘haratha’\\
‘ārocehi’ → ‘ārocetha’

\ifhandbookedition
\clearpage
\fi

\textbf{(b)} Informing the Sangha of the sick bhikkhu's consent:

‘Āyasmā bhante ‘uttaro’ mayhaṃ chandaṃ adāsi, tassa chando mayā āhaṭo, sādhu bhante saṅgho dhāretu.’\\
‘\emph{Ven. sirs, ‘Uttaro Bhikkhu’ has given his consent to me. I have conveyed
  his consent. It is well, ven. sirs, if the Saṅgha holds it to be so.}’

If the bhikkhu conveying consent is senior to the sick bhikkhu:

‘Āyasmā bhante \emph{uttaro}’ → ‘\emph{Uttaro} bhante bhikkhu’

\subsection{Pārisuddhi + Chanda}

When both purity and consent are conveyed to the Sangha:

‘\emph{Uttaro} bhante bhikkhu gilāno mayhaṃ chandañca pārisuddhiñca adāsi, tassa
chando ca pārisuddhi ca mayā āhaṭā, sādhu bhante saṅgho dhāretu.’

‘\emph{Ven. sirs, ‘Uttaro Bhikkhu’ is sick. He has given his consent and purity
  to me. I have conveyed his consent and purity. It is well, ven. sirs, if the
  Sangha holds it to be so.}’

\clearpage

\subsection{Reciting the Pāṭimokkha in Brief}

If there are four or more bhikkhus at the uposatha, but there is an obstruction
to reciting the Pāṭimokkha in full, it may be recited in brief, abbreviating the
recited text and announcing the remainder as `heard' (\emph{sutā}).

On the occasions when an abbreviated recitation is necessary, it is common
practice to recite the Pubbakicca, Nidāna, followed by the Pārājika rules, and
announce the rest as `heard'.

One may include the Saṅghādisesa and Aniyata rules as well, for a longer
recitation, or if the situation demands it, there is allowance to abbreviate
directly after the Nidāna for a shorter recitation.

If the abbreviation is after the Pārājika rules, the chanter concludes with:

\vspace*{\parskip}

\begin{paritta}
`Uddiṭṭhaṃ kho āyasmanto nidānam.\\
Uddiṭṭhā cattāro pārājikā dhammā.\\
Sutā terasa saṅghādisesā dhammā.\\
Sutā dve aniyatā dhammā.\\
Sutā tiṁsa nissaggiyā pācittiyā dhammā.\\
Sutā dve-navuti pācittiyā dhammā.\\
Sutā cattāro pāṭidesanīyā dhammā.\\
Sutā sekhiyā dhammā.\\
Sutā sattādhikaraṇa-samathā dhammā.

Ettakantassa bhagavato suttāgataṃ sutta-pariyāpannaṃ anvaḍḍha-māsaṃ uddesaṃ āgacchati.
Tattha sabbeh'eva samaggehi sammoda-mānehi avivada-mānehi sikkhitabban'ti.

Bhikkhu-pāṭimokkhaṃ niṭṭhitaṃ.'
\end{paritta}

\suttaRef{Vinaya Mukha Vol 2., p.107}

